An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-30 paper/main.tex
weiter

\noindent
For the MNIST experiments (i) end-to-end precision, (ii) timing, (iii)
inference-only precision reduction, and (iv) AMP with gradient clipping, we use
5-fold cross-validation on the training set with random equal splits. Unless
noted otherwise, figures report fold means, and no test set tuning is
performed. In the end-to-end runs, every floating point tensor appearing inside
the training loop (parameters, activations, gradients) is kept at the target
precision during training and evaluation. The timing experiment reuses these
runs to measure mean per-epoch training time and mean evaluation time. For
inference-only reduction, we first train a baseline model in FP64 and then do
inference on copies cast to FP32 and FP16 without any further optimization (so
no \textit{retraining} on the lower precision).\\

The AMP experiment is run on MNIST with gradient clipping, comparing an AMP
setup (FP16 computations with FP32 master weights) against a pure FP32
baseline. We try different global-norm and per-element value-clipping
thresholds to check stability and report test accuracy and training/inference
times. We also run a one-off rounding experiment on FashionMNIST (batch size
\num{256}, FP32 training): after training, only the stored weights are rounded
to $k$ decimal places ($k{=}1\ldots14$) and inference remains FP32. Inputs and
intermediate computations stay at full precision. We report test accuracy as a
function of $k$ and note the impact on storage.

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) improves deep learning training by using both
FP16 and FP32 data types reducing memory usage and increasing computational
speed. However, FP16's limited dynamic range can cause numerical instability,
especially during backpropagation. To address this, Gradient Clipping is used
to cap the magnitude of gradients (either globally or per element) preventing
extreme updates that could hinder convergence. Together, AMP and gradient
clipping enable efficient, stable training by balancing performance and
numerical precision.

>>>>

\noindent
For the MNIST experiments (i) FP64/FP32/FP16 end-to-end training, (ii) effects
on timing, (iii) inference-only casting, and (iv) AMP with gradient clipping,
we use 5-fold cross-validation on the training set with random equal splits.
Unless noted otherwise, figures report fold means, and no test set tuning is
performed. In the end-to-end runs, every floating point tensor appearing inside
the training loop (parameters, activations, gradients) is kept at the target
precision during training and evaluation. The timing experiment reuses these
runs to measure mean per-epoch training time and mean evaluation time. For
inference-only casting, we first train a baseline model in FP64 and then do
inference on copies cast to FP32 and FP16 without any further optimization (so
no \textit{retraining} on the lower precision). The AMP experiment is run on
MNIST with gradient clipping, comparing an AMP setup (FP16 computations with
FP32 master weights) against a pure FP32 baseline. We try different global-norm
and per-element value-clipping thresholds to check stability and report test
accuracy and training/inference times.\\

\noindent
On the FashionMNIST dataset, experiment (v) is a rounding experiment (batch
size \num{256}, FP32 training) which means that after training only the stored
weights are rounded to $k$ decimal places ($k{=}1\ldots14$) and inference
remains at FP32. Inputs and intermediate computations stay at full precision.
We report test accuracy as a function of $k$ and note the impact on storage.

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
AMP improves deep learning training by using both FP16 and FP32 data types
reducing memory usage and increasing computational speed. However, FP16's
limited dynamic range can cause numerical instability, especially during
backpropagation. To address this, Gradient Clipping is used to cap the
magnitude of gradients (either globally or per element) preventing extreme
updates that could hinder convergence. Together, AMP and gradient clipping
enable efficient, stable training by balancing performance and numerical
precision.
